Stateless Model Checking with Data-Race Preemption Points

### 2015-10-20 pldi-16/di.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

% TODO: get latex to put this on page 2 somehow
\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{trees.pdf}
	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
	\caption{Iterative Deepening example.
		The minimal state space (top) includes only mandatory thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
}
	% TODO: reference this label in first few paragraphs of design
	\label{fig:id}
\end{figure}
\section{Design}

TODO

Unlike in chess AI, however, adding more preemption points to a completed state space is not necessarily a linear process, as  % blah blah multiple combinations

% TODO: Introduce abbreviation for SSS-MC here

% define minimal state space and maximal state space
{\bf Terminology.}
Importantly, we distinguish data-race {\em candidates} and data-race {\em bugs}.
Because data-race analysis is false-positive-prone, Landslide reports unprotected access pairs separately from the problems listed above, calling the pair a {\em data race candidate}.
Should a future interleaving, preempting during those accesses, cause a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
If they cannot be reordered at all, it is a {\em false positive}.
For brevity, we count benign data races as a subset of false positives.

\subsection{Verifying data races as buggy or benign}

@@

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for data-race analysis~\cite{eraser}.
% optional: cite tsan above
\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation \cite{estimation}, and data-race detection.
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) similarly to Fair-Bounded Search \cite{bpor}. % TODO: devote a paragraph to this, after talking about drs in locks
% this line can be cut if space is needed
It can test both user- and kernel-level code, though the latter requires user annotations and is limited to timer-driven nondeterminism.
Its bug-detection metrics include assertion failures, deadlock, segfaults, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock checker.

@@

\subsection{Managing multiple state spaces}

@@

>>>>

\begin{figure}[t]
	\includegraphics[width=0.48\textwidth]{trees.pdf}
	%\includegraphics[width=0.15\textwidth]{tree0.pdf}
	%\includegraphics[width=0.225\textwidth]{tree1.pdf}
	%\includegraphics[width=0.25\textwidth]{tree2.pdf}
	%\includegraphics[width=0.35\textwidth]{tree3.pdf}
	\caption{Iterative Deepening example.
		The minimal state space (top) includes only voluntary thread switches, such as {\tt yield()} or {\tt cond\_wait()}~\cite{landslide}.
		Multiple further tests can be run: preempting on calls to {\tt mutex\_lock} alone (left), {\tt mutex\_unlock} alone (right), or both together (bottom).
Each option increases the state space size by an unpredictable factor, so multiple state spaces should be tested in parallel.
Estimation techniques~\cite{estimation} inform which state spaces to prioritize.
}
	\label{fig:id}
\end{figure}
\section{Design}

Motivated by state-of-the-art tools' inflexibility to change their preemption points,
Iterative Deepening searches among different combinations of such points during a test,
deciding on-the-fly whether to pursue each resulting state space or abandon it in favour of smaller ones.
Different state spaces are generated based on {\em subsets} of the preemption points which prior work would use; for example, ``preempt on all calls to {\tt mutex\_lock} but not on {\tt mutex\_unlock}.
We show a visualization in Figure~\ref{fig:id}.

Iterative Deepening depends heavily on state-space estimation \cite{estimation}
to understand which state spaces are likely to complete on time,
in advance of actually testing each interleaving within.
The algorithm's purpose is to make decisions automatically about when to defer exploration of a state space,
so an inexpert user need only provide their total CPU budget as an argument.

Note that Iterative Deepening is a {\em wrapper} algorithm around stateless model checking.
A model checking tool is still used to test each state space, and other reduction techniques are still applicable.
Moreover, because Iterative Deepening treats the set of preemption points as mutable,
it grants the potential to add new ones reactively based on any runtime analysis.
In this paper we focus on dynamic data-race detection~\cite{tsan} as the mechanism for finding new preemption candidates.

\subsection{Terminology}

For the remainder of the paper, we will abbreviate {\em preemption point} (PP),
{\em model checking} (MC),
and {\em single-state-space MC} (SSS-MC) (i.e., the approach of prior work).

Please note the difference between data-race {\em candidates} and data-race {\em bugs}.
Because data-race analysis is false-positive-prone,
we classify unprotected access pairs separately from concrete observable failures, % "concrete"?
calling such pairs {\em data race candidates}.
Should a future interleaving, preempting during those accesses, cause a failure (e.g. assertion or deadlock), then we report a {\em data-race bug}.
Otherwise, if the access pair can be reordered, but does not produce a failure under any interleaving, it is a {\em benign data race}.
If they cannot be reordered at all, it is a {\em false positive}.
For brevity, we count benign data races as a subset of false positives.

We also identify the {\em minimal} and {\em maximal state space} for each test.
The {\em minimal state space} includes only thread switches arising from normal execution (Figure~\ref{fig:id}).
The {\em maximal state space} is simply the one explored by SSS-MC: all statically-available PPs are enabled.
However, should new data-race PPs be added during a test, the new maximal state space will be the one including those as well.

\subsection{Verifying data races as buggy or benign} % FIXME: dont say 'benign'; also, do you have enough left to say that this can even still be a full subsection?

\subsection{Choosing the best job}

% TODO: should work block / find work algorthm here

@@

We chose \landslide~\cite{landslide} as our stateless model checker due to its ability to trace program execution at the granularity of individual instructions and memory accesses, which is mandatory for dynamic data-race analysis.
\landslide~features Dynamic Partial Order Reduction (DPOR) \cite{dpor}, state space estimation, and hybrid lockset/happens-before data-race detection.
It avoids state space cycles (e.g. ad-hoc synchronization with {\tt yield()} or even {\tt xchg} loops) with a heuristic similar to Fair-Bounded Search \cite{bpor}. % TODO: devote a paragraph to this, after talking about drs in locks
% this line can be cut if space is needed
It can test both user- and kernel-level code, although is limited to timer-driven nondeterminism.
Its bug-detection metrics include assertion failure, deadlock, segfault, heap checking (like Valgrind~\cite{valgrind}), and a heuristic infinite loop/livelock check.

@@

\subsection{Quicksand} %Managing multiple state spaces

\quicksand~is an independent program that wraps the execution of several \landslide~instances.
The implementation is roughly 3000 lines of C code.

@@

{\bf Communication protocol.}
The interface to \landslide~, which any similar MC could implement, is in two parts.
First, when starting each job, it writes a configuration file declaring which PPs to use,
% can lose this line due to space
among other options such as mutex-testing mode,
passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the \landslide~process via message-passing on a FIFO pipe.
\landslide~sends a message after each new interleaving is tested, to report updated progress and ETA,
whenever a new data-race candidate is found, and whenever a bug is found.
\quicksand~in turn replies whether the test should suspend/resume due to too high ETA, or quit due to timeout.

{\bf Heuristics.}

### 2015-10-20 pldi-16/evaluation.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

@@

{\bf Partial verification.}
We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-id}}).
While we guarantee safety when preempted on certain combinations of PPs,
ICB ~\cite{chess} guarantees safety under no more than a certain number of preemptions.
%according to the maximum bound reached in the time limit.
We imagine these two guarantees could be each be useful to developers in different scenarios,
and are presently working to combine the two approaches to provide both at once.
One benefit of our technique is that {\tt within\_function}-based Iterative Deepening (discussed above)
would enable expert developers to configure custom subsets of PPs they are most interested in verifying,
according to which modules of a codebase they wish to test.

>>>>

%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
% (as well as on the mutex API boundaries).
% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.

@@

{\bf Partial verification.}
We are not the first to provide a partial verification guarantee when timing out on too-large state spaces (\sect{\ref{sec:eval-id}}).
While we guarantee safety when preempted on certain combinations of PPs,
CHESS
guarantees safety under no more than a certain number of preemptions \cite{chess-icb}.
%according to the maximum bound reached in the time limit.
We imagine these two guarantees could be each be useful to developers in different scenarios,
and are presently working to combine the two approaches to provide both at once.
One benefit of our technique is that {\tt within\_function}-based Iterative Deepening (discussed above)
would enable expert developers to configure custom subsets of PPs they are most interested in verifying,
according to which modules of a codebase they wish to test.

### 2015-10-20 pldi-16/intro.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

% blah blah trite opening sentence
%As parallelism becomes ever more important for achieving high performance in modern-day programs,
%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
or verifying their absence,
%by systematically executing a program along as many distinct interleavings as possible,
by forcing a program to execute each distinct interleaving,
capturing and controlling this nondeterminism in a finite state space.
Unfortunately, these state spaces explode exponentially in the size of the input program.
Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
and search ordering strategies such as Iterative Context Bounding \cite{chess} allow bugs to be found sooner in a given exploration should they exist.

% Can I even make a claim this broad to begin with?
However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
% TODO: You repeat the following 1.5 sentence in the related work section; maybe you can eliminate this redundancy to save space?
For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
It provides an additional data-race analysis to report any violations of this model;
however, data-race analyses are prone to report false positives
%and benign races
which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
%
On the other hand, SPIN \cite{spin}
is able to preempt threads around any shared memory access. Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for all but the most rudimentary test inputs.
%
This work shows how to avoid making that tradeoff decision in advance.

@@

We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
Named after the analogous technique in chess AI \cite{iterative-deepening-chess}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.

>>>>

% blah blah trite opening sentence
%As parallelism becomes ever more important for achieving high performance in modern-day programs,
%so too do advanced concurrency testing techniques become important for verifying the correctness of those programs.
Concurrency bugs are notoriously hard to find and reproduce because they only appear in specific thread interleavings, which arise at random during normal program execution.
{\em Stateless model checking} \cite{verisoft} offers a method for finding such bugs,
or verifying their absence,
%by systematically executing a program along as many distinct interleavings as possible,
by forcing a program to execute each distinct interleaving,
capturing and controlling this nondeterminism in a finite state space.
Unfortunately, these state spaces explode exponentially in the size of the input program.
Reduction techniques such as Dynamic Partial Order Reduction \cite{dpor} and Maximal Causality Reduction \cite{mcr} expand the limits of feasible test completion,
and search ordering strategies such as Iterative Context Bounding \cite{chess-icb} allow bugs to be found sooner in a given exploration should they exist.

% Can I even make a claim this broad to begin with?
However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
% TODO: You repeat the following 1.5 sentence in the related work section; maybe you can eliminate this redundancy to save space?
For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
It provides an additional data-race analysis to report any violations of this model;
however, data-race analyses are prone to report false positives
%and benign races
which require annotations or imprecise heuristics to reduce \cite{racerx,tsan,datacollider}.
%
On the other hand, SPIN \cite{spin}
is able to preempt threads around any shared memory access. Such fine granularity would automatically check if each data race is a real bug, but makes full state space completion intractable for even modestly-sized tests.
%
This work shows how to avoid making that tradeoff decision in advance.

@@

We present \quicksand, a framework for {\em Iterative Deepening} of preemption points during stateless model checking.
Named after the analogous technique in chess AI \cite{iterative-deepening-chess-ai}, our approach likewise makes progressively deeper searches of the state space until a given CPU budget is exhausted.
Rather than attempting to search a single state space with every available preemption point enabled (e.g., preempting on every pthread API call),
\quicksand~tests many different state spaces corresponding to subsets of those points, managing a model checker instance to explore each one.
It estimates the size of each state space to decide when long-running instances should be suspended, and dynamically generates new state spaces based on data race analysis.

### 2015-10-20 pldi-16/related.tex
write 1/2 of design/terminology; move malloc-recycle discussion to impl section

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs.
However, the resulting state space explosion demands that any such tool either
choose a small subset of instructions to consider as PPs.
or be limited to very small test inputs.
%However, every such prior tool we know of has serious drawbacks.
{\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
However, it cannot practically test full-size programs, and requires code to be written in the PROMELA DSL.
%so cannot check implementations directly.

% TODO is it ok to abbreviate ICB and MCR, given we defined the abbrs all the way back in the intro?
{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, MCR \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
These are all orthogonal to our technique.
\textsc{Parrot} \cite{parrot} combines MC with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
Finally, ICB \cite{chess} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

>>>>

{\bf Preempting at instruction granularity} is a prerequisite for using data-race PPs.
However, the resulting state space explosion demands that any such tool either
choose a small subset of instructions to consider as PPs.
or be limited to very small test inputs.
%However, every such prior tool we know of has serious drawbacks.
{\textsc SKI} \cite{ski} approaches kernel code by choosing in advance a random set of instruction offsets from the start of the test,
which is more similar to stress testing or fuzzing than to exhaustive state space exploration.
SPIN \cite{spin} specializes in verifying synchronization primitive implementations such as RCU, which is very similar to our \mxtest~experiment.
However, SPIN explicitly stores visited program states rather than using DPOR, limiting the size of programs that can be practically tested,
and requires code to be written in the PROMELA DSL.
%so cannot check implementations directly.

{\bf Other techniques.} Various improvements to DPOR have been proposed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr}, and DPOR for TSO/PSO \cite{tsopso}.
These are all orthogonal to our technique.
\textsc{Parrot} \cite{parrot} combines MC with a partially-determinizing runtime environment, but still, fewer than half the non-trivial state spaces in their evaluation could be completed, providing a strong argument for \quicksand.
Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to Iterative Deepening, as both approaches provide a concrete partial verification on some subset of interleavings when full completion is intractable (\sect{\ref{sec:future}}).
However, ICB still limits itself to a fixed set of PPs, and so far no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.
